Markov Chain Monte Carlo (MCMC)

### 2017-01-17 bayesian_statistics/MCMC/MCMC.Rmd
multiple testing

---
title: "Markov Chain Monte Carlo (MCMC)"
date: "`r Sys.Date()`"
author: "Ming-Yu Liu"
output:
  rmdformats::readthedown:
    highlight: pygments
---

@@

f <- function(x){
	( sin(.2 * x^3) + sin(x^2) / 5 + 2.5 + sin(10 * x) / 2 ) * dnorm( x, mean = 1 )
}

# boundary for the x axis
x_max <- 6
x_min <- -6

x <- seq( x_min, x_max, length.out = 500 )
y <- f(x)
data <- data.frame( x = x, y = y )

ggplot( data, aes( x, y ) ) + geom_line() + 
geom_ribbon( aes( ymin = 0, ymax = y ), alpha = 0.2 )

@@

# throw N darts (uniformly distributed x-y pairs) at the plot above
N <- 10000
ceiling <- 1.5
xs <- runif( N, min = x_min, max = x_max )
ys <- runif( N, min = 0, max = ceiling )

@@

ggplot( data.frame( x = accepted_samples ) ) + 
geom_histogram( aes( x = x, y = ..density.. ) ) + 
geom_line( data = data, aes( x, y ) )

# study the samples however you want
# mean(accepted_samples)
# quantile(accepted_samples, probs = seq(0, 1, length = 4 ) )

@@

# plot the accepted and unaccepted point
ggplot( df, aes( xs, ys ) ) + 
geom_point( aes( color = accept ), size = 0.5 ) + 
geom_line( data = data, aes( x, y ) )

@@

## MCMC - Metropolis–Hastings Algorithm

@@

lik <- function( x, y ){
	dnorm(x - 3) * dnorm(y - x + 2)
}

grid_values <- seq( x_min, x_max, length = 500 )
grid <- expand.grid( x = grid_values, y = grid_values )
z <- lik( grid$x, grid$y )

# prior probability plot
plot <- ggplot( data = grid, aes( x = x, y = y ) ) + 
		geom_raster( aes( fill = z ) ) + 
		scale_fill_gradient2() + coord_equal() +
		theme_tufte()
plot

@@

# Metropolis-Hastings algorithm (will elaborate on this later)
MCMC <- function( iter, burnin = 0, thin = 1 )
{
	# the list of the first argument for dimnames is to specify the row names,
	# sampling starts at a random point, here 0, 0 
	samples <- matrix( NA, nrow = iter, ncol = 2, 
					   dimnames = list( NULL, c( "x", "y" ) ) )
	samples[ 1, ] <- c( 0, 0 )

	for(i in 2:iter){

		sample <- samples[ i - 1, ]
		for(j in 1:thin){

			# propose a new sample point		
			proposal <- sample + rnorm( n = 2, mean = 0, sd = 1 )

@@

			# flip a coin and accept the 
			# new proposal with probability min( ratio, 1 ),
			# if you don't accept the proposal,
			# then just keep what you had in the last step,
			# meaning nothing changed.
			# note that 1 is evaluated as TRUE
			if( rbinom( 1, size = 1, prob = min( ratios, 1 ) ) )
				sample <- proposal
		}
		samples[ i, ] <- sample
	}
	return( data.frame( samples[ (burnin + 1):iter, ] )	)
}

TracePlot <- function( plot, samples ){
	plot + 
	geom_path( data = samples, aes( x, y ), color = "orange" ) + 
	geom_point( data = samples, aes( x, y ), size = 1 )
}

@@

# plot that shows the first few sampling points are not good ones
samples <- MCMC(iter = 50)
TracePlot( plot, samples )

# plot with burnin, discard the bad sampling points in the beginning
samples <- MCMC( iter = 250, burnin = 125 )
TracePlot( plot, samples )

# use thin to avoid correlation and 
# explore more of the sample space
samples <- MCMC( iter = 250, burnin = 125, thin = 2 )
TracePlot( plot, samples )

@@

@@

1. [The First Stop in Monte Carlo (Methods): Rejection Sampling](http://thexbar.me/2014/11/07/reject-sample/)
2. [mcmc tutorial](https://github.com/davharris/mcmc-tutorial)
3. [Probabilistic Programming and Bayesian Methods for Hackers Chapter 3](http://nbviewer.jupyter.org/github/CamDavidsonPilon/Probabilistic-Programming-and-Bayesian-Methods-for-Hackers/blob/master/Chapter3_MCMC/Chapter3.ipynb)

sessionInfo()

>>>>

---
title: "Markov Chain Monte Carlo (MCMC)"
date: "`r Sys.Date()`"
author: "Ethen Liu"
output:
  rmdformats::readthedown:
    highlight: pygments
---

@@

f <- function(x) {
    ( sin(0.2 * x ^ 3) + sin(x ^ 2) / 5 + 2.5 + sin(10 * x) / 2 ) * dnorm(x, mean = 1)
}

# # boundary for the x axis
x_max <- 6
x_min <- -6

# for simply plotting the function
ggplot( data.frame( x = c(x_max, x_min) ), aes(x) ) + 
stat_function(fun = f)

x <- seq(x_min, x_max, length.out = 500)
y <- f(x)
data <- data.frame(x = x, y = y)
ggplot( data, aes(x, y) ) + geom_line() + 
geom_ribbon( aes(ymin = 0, ymax = y), alpha = 0.2 )

@@

# throw N darts (uniformly distributed x-y pairs) at the plot above
N <- 10000
ceiling <- 1.5
xs <- runif(N, min = x_min, max = x_max)
ys <- runif(N, min = 0, max = ceiling)

@@

ggplot( data.frame(x = accepted_samples) ) + 
geom_histogram( aes(x = x, y = ..density..) ) + 
geom_line( data = data, aes(x, y) )

# we can then study the samples however we want
# mean(accepted_samples)
# quantile(accepted_samples, probs = seq(0, 1, length = 4) )

@@

# plot the accepted and unaccepted point
ggplot( df, aes(xs, ys) ) + 
geom_point( aes(color = accept), size = 0.5 ) + 
geom_line( data = data, aes(x, y) )

@@

## MCMC - Markov chain Monte Carlo

@@

lik <- function(x, y) {
	dnorm(x - 3) * dnorm(y - x + 2)
}

grid_values <- seq(x_min, x_max, length = 500)
grid <- expand.grid(x = grid_values, y = grid_values)
z <- lik(grid$x, grid$y)

# prior probability plot
plot <- ggplot( data = grid, aes(x = x, y = y) ) + 
		geom_raster( aes(fill = z) ) + 
		scale_fill_gradient2() + coord_equal() +
		theme_tufte()
plot

@@

MCMC <- function(iter, burnin = 0, thin = 1) {
	# the list of the first argument for dimnames is to specify the row names,
	# sampling starts at a random point, here 0, 0 
	samples <- matrix( NA, nrow = iter, ncol = 2, 
					   dimnames = list( NULL, c("x", "y") ) )
	samples[1, ] <- c(0, 0)

	for(i in 2:iter) {

		sample <- samples[i - 1, ]	
		for(j in 1:thin) {

			# propose a new sample point (will elaborate on this later)				
			proposal <- sample + rnorm(n = 2, mean = 0, sd = 1)

@@

			# flip a coin and accept the 
			# new proposal with probability min( ratio, 1 ),
			# if you don't accept the proposal,
			# then just keep what you had in the last step,
			# meaning nothing changed.
			# note that 1 is evaluated as TRUE
			if( rbinom( 1, size = 1, prob = min(ratios, 1) ) )
				sample <- proposal
		}
		samples[i, ] <- sample
	}
	return( data.frame(samples[ (burnin + 1):iter, ]) )
}

plot_trace <- function(plot, samples) {
	plot + 
	geom_path( data = samples, aes(x, y), color = "orange" ) + 
	geom_point( data = samples, aes(x, y), size = 1 )
}

@@

# plot that shows the first few sampling points are not good ones
samples <- MCMC(iter = 50)
plot_trace(plot, samples)

# plot with burnin, discard the bad sampling points in the beginning
samples <- MCMC(iter = 250, burnin = 125)
plot_trace(plot, samples)

# use thinning to avoid correlation and
# explore more of the sample space
samples <- MCMC(iter = 250, burnin = 125, thin = 2)
plot_trace(plot, samples)
```

@@

```{r}
sessionInfo()
```

@@

- [Blog: The First Stop in Monte Carlo (Methods): Rejection Sampling](http://thexbar.me/2014/11/07/reject-sample/)
- [Github: mcmc tutorial](https://github.com/davharris/mcmc-tutorial)
- [Notebook: Probabilistic Programming and Bayesian Methods for Hackers Chapter 3](http://nbviewer.jupyter.org/github/CamDavidsonPilon/Probabilistic-Programming-and-Bayesian-Methods-for-Hackers/blob/master/Chapter3_MCMC/Chapter3.ipynb)
